A Gaussian process model of human electrocorticographic data

### 2018-10-12 paper/main.tex
finished a full pass through the main text

\section*{Results}
We used a cross-validation approach to test the accuracy with which
the SuperEEG algorithm reconstructs activity throughout the brain.
For each patient in turn, we estimated full-brain correlation matrices
(Eqn.~\ref{eqn:corrmat}) using data from all of the \textit{other}
patients. This step ensured that the data we were reconstructing could
not also be used to estimate the between-location correlations that
drove the reconstructions via Equation~\ref{eqn:reconstruction}
(otherwise the analysis would be circular).  For that held-out
patient, for each of their electrodes in turn, we used
Equation~\ref{eqn:reconstruction} to reconstruct activity at the
held-out electrode location, using the correlation matrix learned from
all other patients' data as $\hat{K}$, and using activity recorded
from the other electrodes from the held-out patient as
$Y_{s, k, \alpha_s}$.  We then asked: how closely did each of the
SuperEEG-estimated recordings at those electrodes match the observed
recordings from those electrodes (i.e., how closely did the estimated
$\hat{Y}_{s, k, \beta_s}$ match the observed $Y_{s, k, \beta_s}$?).

To illustrate our approach, we first examine an individual held-out
raw LFP trace and its associated SuperEEG-derived reconstruction.
Figure~\ref{fig:recon}A displays the observed LFP from the red
electrode in Figure~\ref{fig:methods}A (blue), and its associated
reconstruction (red), during a 5~s time window during one of the
example patient's six recording sessions.  The two traces match
closely ($r = 0.86, p < 10^{-10}$).  Figure~\ref{fig:recon}B displays a
two-dimensional histogram of the actual versus reconstructed voltages
for the entire 14.2 total hours of recordings from the example
electrode (correlation: $r = 0.91, p < 10^{-10}$).  This example confirms
that the SuperEEG algorithm recovers the recordings from this single
electrode well.  Next, we used this general approach to quantify the
algorithm's performance across the full dataset.

@@

As a stricter benchmark, we compared the quality of these
across-participant reconstructions (i.e., computed using a correlation
model learned from other patients' data) to reconstructions generated
using a correlation model trained using the in-patient's data.  In
other words, for this within-patient benchmark analysis we estimated
$\hat{C}_{s}$ (Eqn.~\ref{eqn:subj_corrmat}) for each patient in turn,
using recordings from all of that patient's electrodes except at the
location we were reconstructing.  These within-patient reconstructions
serve as an estimate of how well data from all of the other electrodes
from that single patient may be used to estimate held-out data from
the same patient.  This allows us to ask how much information about
the activity at a given electrode might be inferred through (a) volume
conductance or other sources of ``leakage'' from activity patterns
measured from the patient's other electrodes and (b) across-electrode
correlations learned from that single patient.  As shown in
Figure~\ref{fig:corrmap}A (gray histogram), the distribution of
within-patient correlations was centered well above zero (mean: 0.32;
$t$-test comparing mean of distribution of $z$-transformed
average patient correlation coefficients to 0: $t(66) = 15.16, p < 10^{-10}$).
However, the across-patient correlations were substantially higher
($t$-test comparing average $z$-transformed within versus across
patient electrode correlations: $t(66) = 9.62, p < 10^{-10}$).  This is an
especially conservative test, given that the across-patient SuperEEG
reconstructions exclude (from the correlation matrix estimates) all
data from the patient whose data is being reconstructed.  We repeated
each of these analyses on a second independent dataset and found
similar results (Fig.~\ref{fig:corrmap}B, D; within versus across
reconstruction accuracy: $t(23) = 6.93, p < 10^{-5}$). We also
replicated this result separately for each of the two experiments from
Dataset 2 (Fig.~\perexpcorrmaps).  This overall finding, that
reconstructions of held-out data using correlation models learned from
\textit{other} patient's data yield higher reconstruction accuracy
than correlation models learned from the patient whose data is being
reconstructed, has two important implications.  First, it implies that
distant electrodes provide additional predictive power to the data
reconstructions beyond the information contained solely in nearby
electrodes.  (This follows from the fact that each patient's
electrodes are implanted in a unique set of locations, so for any
given electrode the closest electrodes in the full dataset are likely
to come from the same patient.)  Second, it implies that the spatial
correlations learned using the SuperEEG algorithm are, to some extent,
similar across people.

@@

The recordings we analyzed from Dataset 1 comprised data collected as
the patients performed a variety of (largely idiosyncratic) tasks
throughout each day's recording session.  That we observed reliable
reconstruction accuracy across patients suggests that the spatial
correlations derived from the SuperEEG algorithm are, to some extent,
similar across tasks.  We tested this finding more explicitly using
Dataset 2.  In Dataset 2, the recordings were limited to times when
each patient was participating in each of two experiments (Experiment
1, a random-word list free recall task, and Experiment 2, a
categorized list free recall task).  We wondered whether a correlation
model learned from data from one experiment might yield good
predictions of data from the other experiment.  Further, we wondered
about the extent to which it might be beneficial or harmful to combine
data across tasks.

To test the task-specificity of the SuperEEG-derived correlation
models, we repeated the above within- and across-patient cross validation
procedures separately for Experiment 1 and Experiment 2 data from
Dataset 2.  We then compared the reconstruction accuracies for held-out
electrodes, for models trained within versus across the two
experiments, or combining across both experiments
(Fig.~\perexptaskrecon).  In every case we found that across-patient
models trained using data from all other patients out-performed
within-patient models trained on data only from the subject
contributing the given electrode ($t$s$(24) > 6.50, p$s$ < 10^{-5}$).  All
reconstruction accuracies also reliably exceeded chance performance
($t$s$(24) > 8.00, p$s$ < 10^{-8}$).  Average reconstruction accuracy was
highest for the across-patient models limited to data from the same
experiment (mean accuracy: $0.68$); next-highest for the
across-patient models that combined data across both experiments (mean
accuracy: $0.61$); and lowest for models trained across task (mean
accuracy: $0.47$).  This result also held for each of the Dataset 2
experiments individually (Fig.~\perexptaskreconseparated).  Taken
together, these results indicate that there are reliable commonalities
in the spatial correlations of full-brain activity across tasks, but
that there are also reliable differences in these spatial correlations
across tasks.  Whereas reconstruction accuracy benefits from
incorporating data from other patients, reconstruction accuracy is
highest when constrained to within-task data, or data that includes a
variety of tasks (e.g., Dataset 1, or combining across the two Dataset
2 experiments).

@@

To gain insights into whether particular electrode locations might be
especially informative, we first computed the average reconstruction
accuracy across all of each patient's electrodes (using the
across-patients cross validation test; black histograms in
Fig.~\ref{fig:corrmap}A and B).  We labeled each patient's electrodes
in each dataset with the average reconstruction accuracy for that
patient.  In other words, we assigned every electrode from each given
patient the same value, reflecting how well the activity patterns at
those electrodes were reconstructed on average.  Next, for each voxel
in the 4~mm$^3$ MNI brain, we computed the average value across any
electrode (from any patient) that came within 20 MNI units of that
voxel's center.  Effectively, we computed an \textit{information
  score} for each voxel, reflecting the average reconstruction
accuracy across any patients with electrodes near each voxel-- where
the averages were weighted to reflect patients who had more electrodes
implanted near that location. This yielded a single map for each
dataset, highlighting regions that are potentially promising
implantation targets in terms of providing full-brain activity
information via SuperEEG (Fig.~\ref{fig:informap}A, B).  Despite task
and patient differences across the two datasets, we nonetheless found
that the maps of the most promising implantation targets derived from
both datasets were similar (voxelwise correlation between information
scores across the two datasets: $r = 0.03, p < 10^{-10}$).  While the
correspondence between the two maps was not perfect, our finding that
there were some commonalities between the two maps lends support to
the notion that different brain areas are differently informative
about full-brain activity patterns.  Further, we found the
intersection of the top 10\% of the most informative voxels between
the two datasets (Fig.~\intersectmap) and found gray matter regions like the right postcentral gyrus, right supramarginal gyrus,
and the right nucleus accumbens that exhibited strong information scores in both datasets may
be especially promising implantation targets.

@@

Here we have taken a model-based approach to studying whether high
spatiotemporal resolution activity patterns throughout the human brain
may be explained by a static connectome model that is shared across
people and tasks.  Specifically, we trained a model to take in
recordings from a subset of brain locations, and then predicted
activity patterns during the same interval, but at \textit{other}
locations that were held out from the model.  Our model, based on
Gaussian process regression, was built on three general hypotheses
about the nature of the correlational structure of neural activity
(each of which we tested).  First, we hypothesized that functional
correlations are stable over time and across tasks.  We found that,
although aspects of the patients' functional correlations that were
stable across tasks, we achieved better reconstruction accuracy when
we trained the model on within-task data~\citep[we acknowledge that
our general approach could potentially be extended to better model
across-task changes, following][and others]{ColeEtal16, TavoEtal16}.
Second, we hypothesized that some of the correlational structure of
people's brain activity is similar across individuals.  Consistent
with this hypothesis, our model explained the data best when we
trained the correlation model using data from \textit{other}
patients-- even when compared to a correlation model trained on the
same patient's data.  Third, we resolved ambiguities in the data by
hypothesizing that neural activity from nearby sources will tend to be
similar, all else being equal.  This hypothesis was supported through
our finding that all of the models we trained that incorporated this
spatial smoothness assumption predicted held-out data well above chance.

One potential limitation of our approach is that it does not provide a
natural means of estimating the precise timing of single-neuron action
potentials.  Prior work has shown that gamma band and broadband
activity in the LFP may be used to estimate the firing rates of
neurons that underly the population contributing to the
LFP~\cite{MillEtal08, MannEtal09, JacoEtal10, CronEtal11}.  Because
SuperEEG reconstructs LFPs throughout the brain, one could in
principle use gamma or broadband power in the reconstructed signals to
estimate the corresponding firing rates (though not the timings of
individual action potentials).

Beyond providing a means of estimating ongoing activity throughout the
brain using already implanted electrodes, our work also has
implications for where to place the electrodes in the first place.
Electrodes are typically implanted to maximize coverage of suspected
epileptogenic tissue.  However, our findings suggest that this
approach could be further optimized.  Specifically, one could leverage
not only the non-invasive recordings taken during an initial
monitoring period (as is currently done routinely), but also
recordings collected from other patients.  We could then ask: given
what we learn from other patients' data (and potentially from the
scalp EEG recordings of this new patient), where should we place a
fixed number of electrodes to maximize our ability to map seizure
foci?  As shown in Figure~\ref{fig:informap}, recordings from
different locations are differently informative in terms of
reconstructing the spatiotemporal activity patterns throughout the
brain.  This property might be leveraged in decisions about where to
surgically implant electrodes in future patients.

@@

\section*{Code availability}
We have published an open-source toolbox implementing the SuperEEG
algorithm.  It may be downloaded
\href{https://supereeg.readthedocs.io/en/latest/}{\underline{here}}. Additionally,
we have provided notebooks for all analyses and figures reported
here.

\section*{Data availability}
The dataset analyzed in this study was generously shared by Michael Kahana.  A portion of Dataset 1 may be downloaded
\href{http://memory.psych.upenn.edu/Request_EEG_access?paper=SedeEtal03}{\underline{here}}.
Dataset 2 may be downloaded
\href{http://memory.psych.upenn.edu/Request_EEG_access?paper=EzzyEtal17}{\underline{here}}.

\section*{Acknowledgements}
We are grateful for useful discussions with Luke J. Chang, Uri Hasson,
Josh Jacobs, Michael Kahana, and Matthijs van der Meer.  We are also
grateful to Michael Kahana for generously sharing the ECoG data we
analyzed in our paper, which was collected under NIMH grant MH55687
and DARPA RAM Cooperative Agreement N66001-14-2-4-032, both to MJK.
Our work was also supported in part by NSF EPSCoR Award Number 1632738
and by a sub-award of DARPA RAM Cooperative Agreement
N66001-14-2-4-032 to JRM.  The content is solely the responsibility of
the authors and does not necessarily represent the official views of
our supporting organizations.

>>>>

\section*{Results}
We used a cross-validation approach to test the accuracy with which
the SuperEEG algorithm reconstructs activity throughout the brain.
For each patient in turn, we estimated full-brain correlation matrices
(Eqn.~\ref{eqn:corrmat}) using data from all of the \textit{other}
patients. This step ensured that the data we were reconstructing could
not also be used to estimate the between-location correlations that
drove the reconstructions via Equation~\ref{eqn:reconstruction}
(otherwise the analysis would be circular).  For that held-out
patient, we held out each electrode in turn.  We used
Equation~\ref{eqn:reconstruction} to reconstruct activity at the
held-out electrode location, using the correlation matrix learned from
all other patients' data as $\hat{K}$, and using activity recorded
from the other electrodes from the held-out patient as
$Y_{s, k, \alpha_s}$.  We then asked: how closely did each of the
SuperEEG-estimated recordings at those electrodes match the observed
recordings from those electrodes (i.e., how closely did the estimated
$\hat{Y}_{s, k, \beta_s}$ match the observed $Y_{s, k, \beta_s}$)?

To illustrate our approach, we first examine an individual held-out
raw LFP trace and its associated SuperEEG-derived reconstruction.
Figure~\ref{fig:recon}A displays the observed LFP from the red
electrode in Figure~\ref{fig:methods}A (blue), and its associated
reconstruction (red), during the 5~s time window during one of the
example patient's six recording sessions shown in
Figure~\ref{fig:methods}E.  The two traces match closely
($r = 0.86, p < 10^{-10}$).  Figure~\ref{fig:recon}B displays a
two-dimensional histogram of the actual versus reconstructed voltages
for the entire 14.2 total hours of recordings from the example
electrode (correlation: $r = 0.91, p < 10^{-10}$).  This example
confirms that the SuperEEG algorithm recovers the recordings from this
single electrode well.  Next, we used this general approach to
quantify the algorithm's performance across the full dataset.

@@

As a stricter benchmark, we compared the quality of these
across-participant reconstructions (i.e., computed using a correlation
model learned from other patients' data) to reconstructions generated
using a correlation model trained using the in-patient's data.  In
other words, for this within-patient benchmark analysis we estimated
$\hat{C}_{s}$ (Eqn.~\ref{eqn:subj_corrmat}) for each patient in turn,
using recordings from all of that patient's electrodes except at the
location we were reconstructing.  These within-patient reconstructions
serve as an estimate of how well data from all of the other electrodes
from that single patient may be used to estimate held-out data from
the same patient.  This allows us to ask how much information about
the activity at a given electrode might be inferred through (a) volume
conductance or other sources of ``leakage'' from activity patterns
measured from the patient's other electrodes and (b) across-electrode
correlations learned from that single patient.  As shown in
Figure~\ref{fig:corrmap}A (gray histogram), the distribution of
within-patient correlations was centered well above zero (mean: 0.32;
$t$-test comparing mean of distribution of $z$-transformed average
patient correlation coefficients to 0: $t(66) = 15.16, p < 10^{-10}$).
However, the across-patient correlations were substantially higher
($t$-test comparing average $z$-transformed within versus across
patient electrode correlations: $t(66) = 9.62, p < 10^{-10}$).  This
is an especially conservative test, given that the across-patient
SuperEEG reconstructions exclude (from the correlation matrix
estimates) all data from the patient whose data is being
reconstructed.  We repeated each of these analyses on a second
independent dataset and found similar results
(Fig.~\ref{fig:corrmap}B, D; within versus across reconstruction
accuracy: $t(23) = 6.93, p < 10^{-5}$). We also replicated this result
separately for each of the two experiments from Dataset 2
(Fig.~\perexpcorrmaps).  This overall finding, that reconstructions of
held-out data using correlation models learned from \textit{other}
patient's data yield higher reconstruction accuracy than correlation
models learned from the patient whose data is being reconstructed, has
two important implications.  First, it implies that distant electrodes
provide additional predictive power to the data reconstructions beyond
the information contained solely in nearby electrodes.  (This follows
from the fact that each patient's grid, strip, and depth electrodes
are implanted in a unique set of locations, so for any given electrode
the closest electrodes in the full dataset tend to come from the
same patient.)  Second, it implies that the spatial correlations
learned using the SuperEEG algorithm are, to some extent, similar
across people.

@@

The recordings we analyzed from Dataset 1 comprised data collected as
the patients performed a variety of (largely idiosyncratic) tasks
throughout each day's recording session.  That we observed reliable
reconstruction accuracy across patients suggests that the spatial
correlations derived from the SuperEEG algorithm are, to some extent,
similar across tasks.  We tested this finding more directly using
Dataset 2.  In Dataset 2, the recordings were limited to times when
each patient was participating in each of two experiments (Experiment
1, a random-word list free recall task, and Experiment 2, a
categorized list free recall task).  We wondered whether a correlation
model learned from data from one experiment might yield good
predictions of data from the other experiment.  Further, we wondered
about the extent to which it might be beneficial or harmful to combine
data across tasks.

To test the task-specificity of the SuperEEG-derived correlation
models, we repeated the above within- and across-patient cross validation
procedures separately for Experiment 1 and Experiment 2 data from
Dataset 2.  We then compared the reconstruction accuracies for held-out
electrodes, for models trained within versus across the two
experiments, or combining across both experiments
(Fig.~\perexptaskrecon).  In every case we found that across-patient
models trained using data from all other patients out-performed
within-patient models trained on data only from the subject
contributing the given electrode ($t$s$(23) > 6.50, p$s$ < 10^{-5}$).  All
reconstruction accuracies also reliably exceeded chance performance
($t$s$(23) > 8.00, p$s$ < 10^{-8}$).  Average reconstruction accuracy was
highest for the across-patient models limited to data from the same
experiment (mean accuracy: $0.68$); next-highest for the
models that combined data across both experiments (mean
accuracy: $0.61$); and lowest for models trained across tasks (mean
accuracy: $0.47$).  This result also held for each of the Dataset 2
experiments individually (Fig.~\perexptaskreconseparated).  Taken
together, these results indicate that there are reliable commonalities
in the spatial correlations of full-brain activity across tasks, but
that there are also reliable differences in these spatial correlations
across tasks.  Whereas reconstruction accuracy benefits from
incorporating data from other patients, reconstruction accuracy is
highest when constrained to within-task data, or data that includes a
variety of tasks (e.g., Dataset 1, or combining across the two Dataset
2 experiments).

@@

To gain insights into whether particular electrode locations might be
especially informative, we first computed the average reconstruction
accuracy across all of each patient's electrodes (using the
across-patients cross validation test; black histograms in
Fig.~\ref{fig:corrmap}A and B).  We labeled each patient's electrodes
in each dataset with the average reconstruction accuracy for that
patient.  In other words, we assigned every electrode from each given
patient the same value, reflecting how well the activity patterns at
those electrodes were reconstructed on average.  Next, for each voxel
in the 4~mm$^3$ MNI brain, we computed the average value across any
electrode (from any patient) that came within 20 MNI units of that
voxel's center.  Effectively, we computed an \textit{information
  score} for each voxel, reflecting the average reconstruction
accuracy across any patients with electrodes near each voxel-- where
the averages were weighted to reflect patients who had more electrodes
implanted near that location. This yielded a single map for each
dataset, highlighting regions that are potentially promising
implantation targets in terms of providing full-brain activity
information via SuperEEG (Fig.~\ref{fig:informap}A, B).  Despite task
and patient differences across the two datasets, we nonetheless found
that the maps of the most promising implantation targets derived from
both datasets were similar (voxelwise correlation between information
scores across the two datasets: $r = 0.20, p < 10^{-10}$).  While the
correspondence between the two maps was imperfect, our finding that
there were some commonalities between the two maps lends support to
the notion that different brain areas are differently informative
about full-brain activity patterns.  We also examined the intersection
between the top 10\% most informative voxels across the two datasets
(white outlines in Fig.~\ref{fig:informap}A, B, Fig.~\intersectmap).
Supporting the notion that structures that are highly interconnected
with the rest of the brain might be especially good targets for
implantation, this intersecting set of voxels with the highest
information scores included major portions of the dorsal attention
network (e.g., inferior parietal lobule, precuneus, inferior temporal
gyrus, thalamus, and striatum) as well as some portions of the default
mode network (e.g., angular gyrus) that are highly interconnected with
a large proportion of the brain's gray matter~\citep[e.g., ][]{TomaVolk11}.

@@

Here we have taken a model-based approach to studying whether high
spatiotemporal resolution activity patterns throughout the human brain
may be explained by a static connectome model that is shared across
people and tasks.  Specifically, we trained a model to take in
recordings from a subset of brain locations, and then predicted
activity patterns during the same interval, but at \textit{other}
locations that were held out from the model.  Our model, based on
Gaussian process regression, was built on three general hypotheses
about the nature of the correlational structure of neural activity
(each of which we tested).  First, we hypothesized that functional
correlations are stable over time and across tasks.  We found that,
although aspects of the patients' functional correlations were
stable across tasks, we achieved better reconstruction accuracy when
we trained the model on within-task data~\citep[we acknowledge that
our general approach could potentially be extended to better model
across-task changes, following][and others]{ColeEtal16, TavoEtal16}.
Second, we hypothesized that some of the correlational structure of
people's brain activity is similar across individuals.  Consistent
with this hypothesis, our model explained the data best when we
trained the correlation model using data from \textit{other}
patients-- even when compared to a correlation model trained on the
same patient's data.  Third, we resolved ambiguities in the data by
hypothesizing that neural activity from nearby sources will tend to be
similar, all else being equal.  This hypothesis was supported through
our finding that all of the models we trained that incorporated this
spatial smoothness assumption predicted held-out data well above chance.

One potential limitation of our approach is that it does not provide a
natural means of estimating the precise timing of single-neuron action
potentials.  Prior work has shown that gamma band and broadband
activity in the LFP may be used to estimate the firing rates of
neurons that underly the population contributing to the
LFP~\cite{MillEtal08, MannEtal09, JacoEtal10b, CronEtal11}.  Because
SuperEEG reconstructs LFPs throughout the brain, one could in
principle use gamma or broadband power in the reconstructed signals to
estimate the corresponding firing rates (though not the timings of
individual action potentials).

Beyond providing a means of estimating ongoing activity throughout the
brain using already implanted electrodes, our work also has
implications for where to place the electrodes in the first place.
Electrodes are typically implanted to maximize coverage of suspected
epileptogenic tissue.  However, our findings suggest that this
approach could be further optimized.  Specifically, one could leverage
not only the non-invasive recordings taken during an initial
monitoring period (as is currently done routinely), but also
recordings collected from other patients.  We could then ask: given
what we learn from other patients' data (and potentially from the
scalp EEG recordings of this new patient), where should we place a
fixed number of electrodes to maximize our ability to map seizure
foci?  As shown in Figures~\ref{fig:informap} and~\intersectmap, recordings from
different locations are differently informative in terms of
reconstructing the spatiotemporal activity patterns throughout the
brain.  This property might be leveraged in decisions about where to
surgically implant electrodes in future patients.

@@

\section*{Code availability}
We have published an open-source toolbox implementing the SuperEEG
algorithm.  It may be downloaded
\href{https://supereeg.readthedocs.io/en/latest/}{\underline{here}}. Additionally,
we have provided code for all analyses and figures reported in the
current manuscript, available
\href{https://github.com/ContextLab/supereeg_paper}{\underline{here}}.

\section*{Data availability}
The dataset analyzed in this study was generously shared by Michael
J. Kahana.  A portion of Dataset 1 may be downloaded
\href{http://memory.psych.upenn.edu/Request_EEG_access?paper=SedeEtal03}{\underline{here}}.
Dataset 2 may be downloaded
\href{http://memory.psych.upenn.edu/Request_EEG_access?paper=EzzyEtal17}{\underline{here}}.

\section*{Acknowledgements}
We are grateful for useful discussions with Luke J. Chang, Uri Hasson,
Josh Jacobs, Michael J. Kahana, and Matthijs van der Meer.  We are
also grateful to Michael J. Kahana for generously sharing the ECoG
data we analyzed in our paper, which was collected under NIMH grant
MH55687 and DARPA RAM Cooperative Agreement N66001-14-2-4-032, both to
M.J.K.  Our work was also supported in part by NSF EPSCoR Award Number
1632738 and by a sub-award of DARPA RAM Cooperative Agreement
N66001-14-2-4-032 to J.R.M.  The content is solely the responsibility
of the authors and does not necessarily represent the official views
of our supporting organizations.
